### 2020-04-01 thesis/content/chapter-ltd.tex
extended 2-wl-gnn definition

@@

@@

>>>>

We have seen that the main difference between 2-\acsp{gnn} and 2-\acs{wl} is their notion of ``neighborhood''.
In this section we describe a novel convolution operator which uses edge tuple neighbors just like 2-\acs{wl} to overcome the limitations of 2-\acsp{gnn}.
This will be done in three steps:
\begin{enumerate*}
	\item We begin by formally defining the 2-\acs{wl} convolution operator.
	\item Then the discriminative and computational power of this operator are analyzed.
	\item Lastly we describe how 2-\acs{wl} convolutions can be implemented efficiently on a modern \ac{gpgpu}.
\end{enumerate*}

@@

Similar to the 2-LWL kernel described in \cref{sec:related:gcr:kernel}, our 2-\acs{wl} convolution operator reduces the computational cost of 2-\acs{wl} via two simplifications:
\begin{enumerate}[label=\textbf{\arabic*.}]
	\item \textbf{2-multisets:}
		Since we assume that graphs are undirected, $e_{ij}$ and $e_{ji}$ have identical feature vectors $x[e_{ij}] = x[e_{ji}] \in \mathcal{X}_{\mathcal{E}}$ and the same 2-\acs{wl} neighborhood.
		Instead of refining/convolving the feature vectors of 2-tuples $(v_i, v_j)$ we can therefore refine/convolve the feature vectors of 2-multisets $\ldblbrace v_i, v_j \rdblbrace$.
		This halves the number of feature vectors without affecting the discriminative or computational power of the 2-\acs{wl} convolution operator.
		To simplify the notation we assume that $e_{ij} = e_{ji} = \ldblbrace v_i, v_j \rdblbrace$ in the rest of this section.
	\item \textbf{Neighborhood localization:}
		Using the 2-multiset simplification, the original 2-\acs{wl} algorithm refines the color of all multisets $e_{ij} \in \mathcal{V}_G^2$ by hashing its current color and the colors of all neighbors ${\{ \ldblbrace e_{il}, e_{lj} \rdblbrace \}}_{v_l \in \mathcal{V}_G}$.
		This means that the time complexity of a single refinement step is $\mathcal{O}(n^3)$ for $n \coloneqq \left| \mathcal{V}_G \right|$ which quickly becomes infeasible for large graphs.

		To address this issue we reduce both the number of colored edges as well as the number of neighbors of each edge.
		This is achieved by only considering the edges that are part of the so-called \textit{$r$-th power of a given graph $G$} where $r \in \mathbb{N}$ is the freely choosable \textit{neighborhood radius}.
		\begin{defn}
			The \textit{$r$-th power of a graph $G$} is defined as
			\begin{align*}
				G^r \coloneqq \left(\mathcal{V}_G, \left\{ e_{ij} \in \mathcal{V}_G^2\, |\, d_{\mathrm{SP}, G}(v_i, v_j) \leq r \right\}\right) % chktex 21
			\end{align*}
			where $d_{\mathrm{SP}, G}(v_i, v_j)$ is the length of the shortest path between $v_i$ and $v_j$ in $G$.
			The distance of a vertex $v_i \in \mathcal{V}_G$ to itself is defined as $d_{\mathrm{SP}, G}(v_i, v_i) \coloneqq 0$.
			Note that $G^1$ does not generally equal $G$ because $G^1$ has self-loop edges $e_{ii} \in \mathcal{E}_{G^1}$ at all vertices.
		\end{defn}
		For the neighborhood radius $r = 1$ only the self-loop edges ${\{ e_{ii} \}}_{v_i \in \mathcal{V}_G}$ and the edges $\mathcal{E}_{G}$ are considered;
		for $r > 1$ edges between indirectly connected vertices are considered as well.
		\Cref{fig:ltd:neighborhood-radius} illustrates this for varying neighborhood radii.
		Through the reduction of the considered edges, the neighbors of each $e_{ij} \in \mathcal{E}_{G^r}$ are in turn reduced to the common $r$-neighbors of $v_i$ and $v_j$, i.e.\ $\{ \ldblbrace e_{il}, e_{lj} \rdblbrace\, |\, v_l \in {\Gamma_{G^r}(v_i) \cap \Gamma_{G^r}(v_j)} \}$. % chktex 21
		\begin{figure}[t]
			\centering
			\includegraphics[width=0.9\linewidth]{gfx/ltd/neighborhood-radius.pdf}
			\caption{
				Illustration of the powers of the six-cycle graph for varying $r$.
				The self-loop edge at each of the vertices is not explicitly shown.
				For $r = 3$ all possible edges between the six vertices will be considered just as in the original 2-\acs{wl} algorithm.
			}\label{fig:ltd:neighborhood-radius}
		\end{figure}

		Let us now consider what the reduced number of considered edges and the reduced number of edge neighbors implies for the runtime of a refinement step.
		If $G$ is a sparse graph with the maximum vertex degree $d \coloneqq \max_{v \in \mathcal{V}_G} \left|\Gamma_G(v)\right|$, the number of considered edges is bounded by $\mathcal{O}(n d^r)$ where each edge has at most $\mathcal{O}(d^r)$ neighbors.
		Consequently the time complexity of a refinement step becomes $\mathcal{O}(n d^{2r})$ which is a significant improvement over the $\mathcal{O}(n^3)$ bound of a full 2-\acs{wl} refinement step (assuming $d \ll n$).
\end{enumerate}
Based on the 2-multiset and the neighborhood localization simplifications we can now define the 2-\ac{wl} convolution operator.
\begin{defn}\label{defn:ltd:wl2-conv-init}
	The \textit{initial feature matrix $Z_G^{(0)}$} of the 2-\acs{wl} convolution operator with the neighborhood radius $r \in \mathbb{N}$ contains both the vertex features $x[v_i] \in \mathcal{X}_{\mathcal{V}} = \mathbb{R}^{d_{\mathcal{V}}}$ as well as the edge features $x[e_{ij}] \in \mathcal{X}_{\mathcal{E}} = \mathbb{R}^{d_{\mathcal{E}}}$ of a given graph $G$.
	More specifically $Z_G^{(0)} \in \mathbb{R}^{{|\mathcal{E}_{G^r}|} \times (d_{\mathcal{V}} + d_{\mathcal{E}})}$ assigns a row vector $Z_G^{(0)}[e_{ij}]$ to all edges $e_{ij} \in \mathcal{E}_{G^r}$.
	Those initial edge feature vectors are defined by the following vector concatenation:
	\begin{align*}
		Z_G^{(0)}[e_{ij}] \coloneqq \left( \begin{cases}
			x[v_i] & \text{if $i = j$} \\
			\mathbf{0} & \text{else}
		\end{cases} \right) \oplus \left( \begin{cases}
			x[e_{ij}] & \text{if $e_{ij} \in \mathcal{E}_G$} \\
			\mathbf{0} & \text{else}
		\end{cases} \right)
	\end{align*}
\end{defn}
\begin{defn}\label{defn:ltd:wl2-conv-step}
	We define the \textit{2-\acs{wl} graph convolution operator} as
	\begin{gather*}
		Z_G^{(t)}[e_{ij}] \coloneqq \sigma\left( Z_G^{(t-1)}[e_{ij}] W_{\mathrm{L}}^{(t)} + \smashoperator[lr]{\sum_{v_l \in {\Gamma_{G^r}(v_i) \cap \Gamma_{G^r}(v_j)}}} \kappa^{(t)}\left( Z_G^{(t-1)}[e_{ij}], \ldblbrace Z_G^{(t-1)}[e_{il}], Z_G^{(t-1)}[e_{lj}] \rdblbrace \right) \right) \\
		\text{with } \kappa^{(t)}(z_{ij}, \ldblbrace z_{il}, z_{lj} \rdblbrace) \coloneqq \left( z_{ij} W_{\mathrm{F}}^{(t)} \right) \odot \sigma_{\Gamma}\left(\left( z_{il} + z_{lj} \right) W_{\Gamma}^{(t)} \right)
		\text{.}
	\end{gather*}
\end{defn}
The convolution operator from \cref{defn:ltd:wl2-conv-step} is parameterized by the three matrices $W_{\mathrm{L}}^{(t)}, W_{\mathrm{F}}^{(t)},  W_{\Gamma}^{(t)} \in \mathbb{R}^{d^{(t-1)} \times d^{(t)}}$ and uses two freely choosable nonlinearities $\sigma$ and $\sigma_{\Gamma}$.
There are three properties which motivate this particular choice of convolution layer:
\begin{enumerate}[label=\textbf{\arabic*.}]
	\item \textbf{Simulation of \acp{mlp}:}
		By choosing $W_{F}^{(t)} = \mathbf{0}$ the convolution behaves like a regular fully connected layer.
		A stack of 2-\acs{wl} convolution layers can therefore simulate arbitrary \acp{mlp}.
	\item \textbf{Perservation of edge pair information:}
		The 2-\acs{wl} convolution layer computes a feature vector for each neighbor $\ldblbrace e_{il}, e_{lj} \rdblbrace$ of $e_{ij}$ via $\sigma_{\Gamma}\left(\left( z_{il} + z_{lj} \right) W_{\Gamma}^{(t)} \right)$.
		A 2-\acs{gnn} use a similar formulation but leaves out the inner nonlinearity $\sigma_{\Gamma}$;
		as we saw in \cref{fig:ltd:2gnn-2wl-diff} this causes 2-\acsp{gnn} to lose the edge pair information due to the commutativity and associativity of $+$.
	\item \textbf{Context-dependent neighborhood filtering:}
		Instead of filtering out all neighbors, $W_{F}^{(t)}$ also allows to filter the neighborhood of $e_{ij}$ more selectively.
		We can interpret $z_{ij} W_{\mathrm{F}}^{(t)} \in \mathbb{R}^{d^{(t)}}$ as a row vector of feature dimension weights.
		The feature dimensions of neighbors are rescaled via those weights which allows the model to aggregate different types of neighbors depending on $z_{ij}$.
\end{enumerate}

@@

In this section we will analyze the discriminative and computational power of the 2-\ac{wl} convolution operator that was just defined.
We begin by showing that 2-\acs{wl}-\acsp{gnn} are at least as powerful as 1-\acs{wl}.

\begin{prop}
	2-\acs{wl}-\acsp{gnn} can simulate arbitrary \acp{gin}.
\end{prop}
